\mytitle

### 2016-10-08 pdocs/thesis/abstract.tex
pdocs and debugDraw for ElasticPlastic2Constraint
Aki's comments on article and in general

In this work suitable methods for handling of plasticity in physics engine are presented.
Handling of plasticity basically means that objects involved may absorb energy and possibly break.

@@

Keywords: Physics engine, simulation, dynamics, plasticity, deformation.

>>>>

In this work suitable methods for handling of plasticity in physics engine are presented.
Handling of plasticity basically means that involved bodies may absorb energy and possibly break.

@@

Keywords: Physics engine, simulation, dynamics, plasticity, inelastic, multibody dynamics,  physical simulation, deformation.

### 2016-10-08 pdocs/thesis/article-conclusion.tex
Aki's comments on article and in general

This study has presented an approach to account plastic deformation in 
velocity based formulation.
In the introduced method, the plastic deformation takes place if force or moment exceeds given 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
Approach presented in this work can be used in gaming industry to provide more realistic 
simulations without significant extra work. 
For gaming purposes presented method works 
best in scenarios where connected parts are relatively heavy. This allows normal 
integration timestep to be used without stability issues.

This kind of metodology also opens large area of combining  structural analysis
methods to realtime simulation frameworks.
Handling of significant axial and shear forces and moments at same time
is one possible area of further studies.
Handling of more complex structures is already being studied by author.

There is also work to be done in the area of performance optimization.
One possible area is reusing once calculated values if memory usage is not highly limited.
This could done e.g. when dealing with high frequency modes.
Another area is usage of graphics processing unit (GPU) for calculation.
\bullet\ already has already experimental GPU pipeline but most constraint types are not 
yet supported. GPU support was one reason for selecting
it to be used in this study

>>>>

This study presents an approach to account plastic deformation in 
velocity based formulation.
In the introduced method, the plastic deformation takes place if force or moment exceeds given 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
Approach presented in this work can be used in gaming industry to provide realistic 
simulations. 
For gaming purposes presented method works 
best in scenarios where connected parts are heavy. This allows normal 
integration timestep to be used without stability issues.

This kind of metodology also opens large area of combining  structural analysis
methods to realtime simulation frameworks.
Handling of significant axial and shear forces and moments at same time
is one possible area of further studies.

There is also work to be done in the area of performance optimization.
One possible area is reusing once calculated values if memory usage is not limited.
This could done e.g. when dealing with high frequency modes.
Another area is usage of graphics processing unit (GPU) for calculation.
\bullet\ already has already experimental GPU pipeline but most constraint types are not 
yet supported. GPU support was one reason for selecting
it to be used in this study

### 2016-10-08 pdocs/thesis/article-introduction.tex
Aki's comments on article and in general

Plasticity is not typically taken into account in gaming solutions. 
Breaking of various objects often takes place based on collision or impulse.
Nevertheless, breaking of steel or reinforced concrete structures using this approach 
is not appropriate making a simulation to look realistic. Theory for handling of plasticity 
has been presented already in \cite{cg1988}. \cite{muller2004point} and \cite{muller2005meshless} 
present a method for modeling and animating of elastic and plastic objects in real-time using 
point based animation. This approach is not been widely used in computer games.  
On major issue is collision handling of deformable objects.

This study will introduce an approach to account plastic deformation in game applications.   
In the introduced method, the plastic deformation takes place if force or moment exceeds predefined 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
The approach is based on using joint motors to model plasticity. 
This study extends a method introduced by
\citet{erleben.thesis} 
%\citet[p.~90]{erleben.thesis} 
which is originally proposed for modelling friction in joints. 
In the introduced method adjacent objects are connected by motors. 
Motor power production limits are estimated based on plastic section modulus. 
Joint breaking is accounted by summing plastic deformation and comparing it to 
predefined material based limit. Elastic part of deformation is modelled by employing 
spring description which is based on modification of existing constraint in \bullet.

Approach presented in this work can be used in gaming industry to provide more realistic 
simulations without significant extra work. For gaming purposes presented method works 
best in scenarios where connected parts are relatively heavy. This allows normal 
integration timestep to be used without stability issues. 
This kind of metodology also opens large area of combining old structural analysis
methods to modern simulation frameworks.

>>>>

Plasticity is not typically taken into account in gaming solutions. 
Breaking of various bodies often takes place based on collision or impulse.
Nevertheless, breaking of steel or reinforced concrete structures using this approach 
is not appropriate making a simulation to look realistic. Theory for handling of plasticity 
has been presented already in \cite{cg1988}. \cite{muller2004point} and \cite{muller2005meshless} 
present a method for modeling and animating of elastic and plastic bodies in real-time using 
point based animation. This approach is not been widely used in computer games.  
On major issue is collision handling of deformable bodies.

This study will introduce an approach to account plastic deformation in game applications.   
In the introduced method, the plastic deformation takes place if force or moment exceeds predefined 
limit, deformation absorbs energy and joint breaks if plastic capacity is exceeded. 
The approach is based on using joint motors to model plasticity. 
This study extends a method introduced by
\citet{erleben.thesis} 
%\citet[p.~90]{erleben.thesis} 
which is originally proposed for modelling friction in joints. 
In the introduced method adjacent bodies are connected by motors. 
Motor power production limits are estimated based on plastic section modulus. 
Joint breaking is accounted by summing plastic deformation and comparing it to 
predefined material based limit. Elastic part of deformation is modelled by employing 
spring description which is based on modification of existing constraint in \bullet.

Approach presented in this work can be used in gaming industry to provide more realistic 
simulations without significant extra work. For gaming purposes presented method works 
best in scenarios where connected parts are heavy. This allows normal 
integration timestep to be used without stability issues. 
This kind of metodology also opens large area of combining old structural analysis
methods to modern simulation frameworks.

### 2016-10-08 pdocs/thesis/article-section-2.tex
Aki's comments on article and in general

where $\vec{r}_{anc}^{\,i}$ is used to define at which point
joint constraint is applied.

@@

Constraint processing in \bullet\ is based on ODE, \cite{ode}.
Mathematical background and detailed examples are available by \cite{ode.joints}.
Joints are also discussed in detail in  
\citet{erleben.thesis}.
%\citet[p.~60-90]{erleben.thesis}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint.

@@

Material plasticity can be accounted in games by using suitable coefficient of restitution.
This provides reasonable means to simulate loss of energy in collisions.
Simulation of breaking of objects made of ductile material can be made more realistic by splitting rigid bodies
to multiple bodies which are connected by energy absorbing joints.
Typical engineering stress-strain curve of ductile steel is shown in Figure \ref{fig:sscurve}.

@@

where $\sigma$ is stress, $E$ is Youngs modulus and $f_y$ is yield stress.
Engineering stress and strain mean that original dimensions are used in stress calculation,
\citet{dowling}.
%\citet[p.~108]{dowling}.
Stress-strain curve is not drawn to scale as elastic strain could not be seen as it is typically 0.001 to 0.005.

In this work elastic-fully plastic material model is used in most scenarios.
Having elastic part allows elastic displacements for slender structures. 
Elastic material behavior is ignored in approach introduced in this work provided
that deformation is related to higher frequency
than integration stability would allow.
It should be noted that geometry
of objects is not updated during analysis and thus engineering stress-strain properties should
be used.

Strain hardening is taken into account in this work mainly by assuming that plasticity in bending
expands, 
\citet{dowling}.
%\citet[p.~672]{dowling}.
Material that starts to yield first is hardened and yielding moves slightly.
This can be seen e.g. by bending paperclip. It does not break at low angles but can take few full bends.

Difference between elastic and plastic section modulus is shown in Figure \ref{fig:wp}.

@@

If stress is below yield limit $f_y$, stress and strain are linear within material.
If cross section is fully plastic, stress is assumed to be at yield level over whole cross section and 
so plastic section modulus is higher than elastic section modulus.

Plasticity is handled by defining maximum forces
in Equations \ref{eq:lambdaLow} and  
\ref{eq:lambdaHigh} using plastic capasities as maximum forces.

@@

Maximum forces and moments for 
rectangular section using constant yield stress
are summarized in Figure \ref{fig:maxForces}.
Yield shear stress is assumed to be $ 0.5\, f_y$ using Tresca yield critetion.
If von Mises yield criterion is used 0.5 is replaced by 0.58 ($1/\sqrt{3}$).
These are not exact values in multiaxial stress state but they
should be quite usable in most cases.

@@

For rotation around y-axis there is closed form solution only for
circular cross sections.
Given approximation is 
best suited for cases where $b$ and $h$ are quite similar.
Better approximation for any given $b$ and $h$ can be obtained 
by integrating distance from center of joint over cross section and
multiplying it with yield shear stress e.g. using octave.
Example of calculation of maximum moment  around $\vec{r}_{anc}^{\,i} $
is shown in Figure \ref{fig:octave-mp}

@@

Basic idea in this work can be tested with any framework having motors and hinge constraints.
This can be done by setting target velocity of motor to zero and limiting maximum motor impulse to plastic moment 
multiplied by timestep.

>>>>

In Figure \ref{fig:eom-joint}, $\vec{r}_{anc}^{\,i}$ is used to define at which point
joint constraint is applied.

@@

Constraint processing in \bullet\ is based on ODE, \cite{ode}.
Joints are also discussed in detail in  
\citet{erleben.thesis}.
%\citet[p.~60-90]{erleben.thesis}.
Equations \ref{eq:constraintEquation}, \ref{eq:lambdaLow} and
\ref{eq:lambdaHigh} 
are created for each constraint. 
Derivation for terms in Equation \ref{eq:constraintEquation}
can be done using position and orientation of connected bodies.
E.g. for ball joint formulation is based on both joint points having same position.
In contact cases formulation is easier if it is done using velocities, \cite{ode.joints}.

@@

Material plasticity can be accounted in games by using suitable coefficient of restitution.
This provides reasonable means to simulate loss of energy in collisions.
Simulation of breaking of bodies made of ductile material can be made more realistic by splitting rigid body
to multiple bodies which are connected by energy absorbing joints.
Typical engineering stress-strain curve of ductile steel is shown in Figure \ref{fig:sscurve}.

@@

In Figure \ref{fig:sscurve}, $\sigma$ is stress, $E$ is Youngs modulus and $f_y$ is yield stress.
Engineering stress and strain mean that original dimensions are used in stress calculation,
\citet{dowling}.
%\citet[p.~108]{dowling}.
Stress-strain curve is not drawn to scale as elastic strain could not be seen as it is typically 0.001 to 0.005.

In this work elastic-fully plastic material model is used in most scenarios.
Having elastic part allows elastic displacements for slender structures. 
Elastic material behavior is ignored in approach introduced in this work provided
that deformation is related to higher frequency
than integration stability would allow.
It should be noted that geometry
of bodies is not updated during analysis and thus engineering stress-strain properties should
be used.

In this work, strain hardening is taken into account by assuming that plasticity in bending
expands, 
\citet{dowling}.
%\citet[p.~672]{dowling}.
Material that starts to yield first is hardened and as a result of which yielding moves.
This can be seen e.g. by bending paperclip. It does not break at low angles but can take few full bends.

Difference between elastic and plastic section modulus is depicted in Figure \ref{fig:wp}.

@@

As shown in Figure 2.4, if stress is below yield limit $f_y$, stress and strain are linear within material.
If cross section is fully plastic, stress is assumed to be at yield level over whole cross section such that 
plastic section modulus is higher than elastic section modulus.

In this work, plasticity is handled by defining maximum forces
in Equations \ref{eq:lambdaLow} and  
\ref{eq:lambdaHigh} using plastic capasities which are defined below.

@@

Maximum forces and moments for 
rectangular section using constant yield stress
are summarized in Figure \ref{fig:maxForces}.
Yield shear stress is assumed to be $ 0.5\, f_y$ using Tresca yield critetion.
If von Mises yield criterion is used 0.5 is replaced by 0.58 ($1/\sqrt{3}$), \cite{dowling}.
% p. 262, p. 268
These are not exact values in multiaxial stress state but they
are usable in most cases.

@@

For rotation around $y$-axis there is closed form solution only for
circular cross sections.
Given approximation is 
best suited for cases where $b$ and $h$ are similar.
Better approximation for any given $b$ and $h$ can be obtained 
by integrating distance from center of joint over cross section and
multiplying it with yield shear stress e.g. using octave, \cite{octave}.
Example of calculation of maximum moment  around $\vec{r}_{anc}^{\,i} $
is shown in Figure \ref{fig:octave-mp}

@@

Basic idea introduced in this study can be tested with any framework having motors and hinge constraints.
This can be done by setting target velocity of motor to zero and limiting maximum motor impulse to plastic moment 
multiplied by timestep.

### 2016-10-08 pdocs/thesis/article-section-3.tex
Aki's comments on article and in general

System has only one dynamic rigid body which is three meters high block of 0.04 \% steel enforced concrete. 
Cross section is one square meter.
Constraint is set so that connecting frame is at top of block. 
In this case frame could be anywhere but if multiple bodies are involved, connecting frame must be defined
so that it reflects scenario under investigation.
Concrete density is $2000\ \frac{kg}{m^3}$ and steel density is $7800\ \frac{kg}{m^3}$. 
Steel is assumed to handle load in elastic-plastic case. Yield stress of steel is 200 MPa.
Gravity is $10\ \frac{m}{s^2}$. Simulation step is $\frac{1}{60}\ s$ and 10 
iterations are done for single step.
Body is 1.5 meters above rigid ground. In unconstrained case body will hit hit ground at t = 0.548 s 
($\sqrt{2 \cdot 1.5 / 10}$).

Single simulation step is performed using following substeps. 
In this work, changes are made only to constraint setup and update actions.
\begin{enumerate}
\item Apply gravity to each non static body. 
This step makes programming easier as otherwise each program should add forces due to gravity in each step. 
In this case, $\vec{f}_{ext}$ in  Equation \ref{eq:eomV2} gets added by $\lbrace{0,-60000,0}\rbrace$.
\item Predict unconstrained motion for each body controlled by physics simulation.
Static and kinematic bodies are not processed in this step.
Prediction is done based on current linear and angular velocities of the body.
\item Predict contacts. In this phase, continuous collision detection is done based on simplified objects. 
Each rigid body is represented by sphere geometry and contact prediction is done if body moves more than given threshold value during simulation step. Continuous collision detection is configured for each body. 
Typical scenario that requires continuous collision detection is fast moving bodies that would otherwise go through walls.
\item Perform discrete collision detection. All overlapping bodies are processed and manifoldPoints are created for
each detected contact at end of simulation step. 
In unconstrained case, contact is detected after $t$=0.55 s if continuous collision detection is not used and manifolds distance gets negative value (-0.058 m).
\item Calculate simulation islands(groups). Bodies that are near each other based on contact prediction or discrete collision detection
 or connected with constraints are grouped in same group.
\item Solve constraints. Both contact and other constraints are processed in this step. 
 This step is subdivided to setup, iterations and finish phases. 
In setup phase constraints are queried for positional errors for calculation of $c$ in
Equation \ref{eq:constraintEquation} and maximum and minimum impulses in Equations \ref{eq:lambdaLow} and 
\ref{eq:lambdaHigh}.
In finish phase constraint forces are calculated if requested and velocities of bodies are updated.
\item Integrate transforms using velocities modified in previous step.
\item Update actions (callbacks) are called. 
In elastic-plastic case, plasticity is summed and equilibrium point of elastic part is updated if maximum force or moment is exceeded.
\item Activation state of bodies is updated. 
To avoid extra calculation bodies are as default put to sleeping state if linear and angular velocities of body are less than
given threshold values (default 0.8 and 1.0) longer than set time limit (2.0).
\end{enumerate}

@@

Equations \ref{eq:lambdaLow} and \ref{eq:lambdaHigh} are not active for fixed case.
For elastic-plastic case maximum impulse is set to product of yield stress, area of steel enforcement and time step (1330).

Method btSequentialImpulseConstraintSolver::solveGroupCacheFriendlySetup
in \bullet\ was used to pick up values for internal variables described in Figure \ref{fig:solVars}.

\begin{figure}[htb!]
\begin{description}
\item[velError] is calculated using velocities and external impulses of connected bodies.  
 \begin{description}
\item[In constraint cases,] main contibutor is bodies relative speed at joint point.
\item[In contact case,] main contributor is bodies relative speed at point of contact. 
Bodies are not allowed to penetrate each other. Restitution increases velError.
If contact is not penetrating velError is reduced by $penetration * timeStep$.
\end{description}
\item[posError] is calculated by constraint. It is significant factor in designing stable constraints.
 \begin{description}
 \item[In fixed case,] value is about 12 times actual position error. Factor 12 is based on time step (60) 
 and default value of error reduction parameter (erp) which has value of 0.2 in this context.
 \item[In elastic plastic case,]  value is set to zero if impulse would be larger than maximum impulse or
spring simulation cannot be done in stable way.
 \item[In contact case,] value is zero if there is no penetration. For penetration cases it is 
$\frac{-penetration\, erp}{timeStep}$. In contact cases default value for erp is 0.8.
 \end{description}
\item[rhs (c)] is calculated by velError jInv + posError jInv
\item[jInv] is calculated using masses and inertias of connected bodies and constraint geometry. 
 \begin{description}
\item[In constraint cases,] it is mass of body (6000).
\item[In contact case,] it varies below mass of body.
\end{description}
\item[Impulse] is impulse applied to body during timestep.
 \begin{description}
\item[In constraint cases,] it is obtained from btJointFeedback structure.
\item[In contact case,] it is obtained by summing applied impulses from active manifolds.
\end{description}
\item[erp] Error reduction parameter (0...1) is used to handle numerical issues e.q. object drifting. 
Setting erp to 1 would in theory eliminate error in one step but in practice value of 0.2 - 0.8 is used in most cases.
\end{description}
\caption{Internal variables used in \bullet\ in constraint solving.}
\label{fig:solVars}
\end{figure}

@@

Six-dof-spring constraint 2 has optional feature to avoid unstability by automatically softening constraint
spring. Feature is activated if current timeStep is too large for spring-mass system simulation.
The feature is trigged based on the equation below 
\begin{equation} \label{eq:frequencyLimited}
\sqrt{ k /  m_{min}} dt > 0.25 
\end{equation}
where $k$ is elastic spring stiffness between bodies,
$m_{min}$ is smaller of connected masses or inertias and $dt$ is integration timestep.

If this feature is active for elastic-plastic constraint 2 constraint  
it does not report positional error but it allows full plastic capacity to be used for correcting
velocity based error instead of force provided by spring.
In this scenario object behaves as depicted in table \ref{tab:ep2BlockValues} i.e. it does not move at all.

@@

\subsection{Charpy impact test}
Simulation of Charpy impact test was used to benchmark approach in more complicated scenario. 
Material is steel and density is 7800 $\frac{kg}{m^{3}}$. Young’s modulus is 200 GPa.
Specimen dimensions are 10x10x55 mm with 2 mm notch in middle which is taken into account in calculations.
Support anvils initially have 40 mm open space between them. Their width is 40 mm. 
Hammer is 0.5 m wide and 0.25 m high, thickness is 0.02 m. Mass is 19.5 kg.
Hammer has 40 mm draft.
If specimen bends about 1.9 radians (108 degrees) it will go between anvils.
Expected energy loss is product of plastic moment of section, hinge angle needed for specimen to go through supports and 
yield stress of specimen. For hinge angle of 1.9 radians and yield stress of 400 MPa expected energy 
loss is 122 J.

@@

\subsection{Demolisher}
Demolisher is  collection of possible scenarios that can be added to games to provide ductile joints between
rigid bodies.  Rigid concrete bodies are connected by steel enforcement.
Concrete density is 2000 $\frac{kg}{m^3}$ and steel density is 7800 $\frac{kg}{m^3}$. 
Steel is assumed to handle load. Yield stress of steel is 200 MPa.
Gravity is 10  $\frac{m}{s^2}$. Simulation step is $\frac{1}{60} $ s and 10 
iterations are done for single step. Multiple simulation steps are done for each render if needed.

Simulation can be done in real time on modest hardware. 
Figure \ref{fig:demolisher} shows three screenshots from simulation.

@@

Vehicle is modelled as single box btRaycastVehicle with four driving wheels. Wheels are rendered for visual feedback.
In left picture, initial state is shown. Gate in front is modelled to be so slender that it has visible elastic deflection.
Gate support is modelled as heavy box.
Bridge has rigid ramps and supports at both ends.
Gravity is applied gradually to avoid collapse of bridge at start of simulation.
In middle picture,  bridge is shown in state where car has driven once over bridge without stopping and
other bodies have taken light damage.
In right picture, some of joints have been broken.

>>>>

In the first example, system under investigation
has only one dynamic rigid body which is three meters high block of 0.04 \% steel enforced concrete. 
Cross section is one square meter.
Constraint is set so that connecting frame is at top of block. 
In this case frame could be anywhere but if multiple bodies are involved, connecting frame must be defined
so that it reflects scenario under investigation.
Concrete density is $2000\ \frac{kg}{m^3}$ and steel density is $7800\ \frac{kg}{m^3}$. 
Steel is assumed to handle load in elastic-plastic case. Yield stress of steel is 200 MPa.
Gravity is $10\ \frac{m}{s^2}$ to keep numbers simple. 
Simulation step is $\frac{1}{60}\ s$ and 10 
iterations are done for single step.
Body is 1.5 meters above rigid ground. In unconstrained case body will hit hit ground at $t$ = 0.548 s 
($\sqrt{2 \cdot 1.5 / 10}$).

Single simulation step is performed using following substeps. 
In this work, changes are made only to constraint setup and update actions.
\begin{enumerate}
\item Apply gravity to each moveable body. 
This step makes programming easier as otherwise each program should add forces due to gravity in each step. 
In this case, $\vec{f}_{ext}$ in  Equation \ref{eq:eomV2} gets added by $\lbrace{0,-60000,0}\rbrace$.
\item Predict unconstrained motion for each body controlled by physics simulation.
Static and kinematic bodies are not processed in this step.
Prediction is done based on current linear and angular velocities of the body.
\item Predict contacts. In this phase, continuous collision detection is done based on simplified bodies. 
Each rigid body is represented by sphere geometry and contact prediction is done if body moves more than given threshold value during simulation step. Continuous collision detection is configured for each body. 
Typical scenario that requires continuous collision detection is fast moving bodies that would otherwise go through walls.
\item Perform discrete collision detection. All overlapping bodies are processed and manifoldPoints are created for
each detected contact at end of simulation step. 
In unconstrained case, contact is detected after $t$=0.55 s if continuous collision detection is not used and manifolds distance gets negative value (-0.058 m).
\item Calculate simulation islands(groups). Bodies that are near each other based on contact prediction or discrete collision detection
 or connected with constraints are grouped in same group.
\item Solve constraints. Both contact and other constraints are processed in this step. 
 This step is subdivided to setup, iterations and finish phases. 
In setup phase constraints are queried for positional errors for calculation of $c$ in
Equation \ref{eq:constraintEquation} and maximum and minimum impulses in Equations \ref{eq:lambdaLow} and 
\ref{eq:lambdaHigh}.
In finish phase constraint forces are calculated if requested and velocities of bodies are updated.
\item Integrate transforms using velocities modified in previous step.
\item Update actions (callbacks) are called. 
In elastic-plastic case, plasticity is summed and equilibrium point of elastic part is updated if maximum force or moment is exceeded.
\item Activation state of bodies is updated. 
To avoid extra calculation bodies are as default put to sleeping state if linear and angular velocities of body are less than
given threshold values (default 0.8 and 1.0) longer than set time limit (2.0).
\end{enumerate}

@@

Equations \ref{eq:lambdaLow} and \ref{eq:lambdaHigh} are not active for fixed case as
there are no upper nor lower limit for forces.
For elastic-plastic case maximum impulse is set to product of yield stress, 
area of steel enforcement and time step (1330).

Method btSequentialImpulseConstraintSolver::solveGroupCacheFriendlySetup
in \bullet\ is used to pick up values for internal variables described in Figure \ref{fig:solVars}.

\begin{figure}[htb!]
\begin{description}
\item[velError] is calculated using velocities and external impulses of connected bodies.  
 \begin{description}
\item[In constraint cases,] main contibutor is bodies relative speed at joint point.
\item[In contact case,] main contributor is bodies relative speed at point of contact. 
Bodies are not allowed to penetrate each other. Restitution increases velError.
If contact is not penetrating velError is reduced by $penetration * timeStep$.
\end{description}
\item[posError] is calculated by constraint. It is significant factor in designing stable constraints.
 \begin{description}
 \item[In fixed case,] value is about 12 times actual position error. Factor 12 is based on time step (60) 
 and default value of error reduction parameter (erp) which has value of 0.2 in this context.
 \item[In elastic plastic case,]  value is set to zero if impulse would be larger than maximum impulse or
spring simulation cannot be done in stable way.
 \item[In contact case,] value is zero if there is no penetration. For penetration cases it is 
$\frac{-penetration\, erp}{timeStep}$. In contact cases default value for erp is 0.8.
 \end{description}
\item[rhs (c)] is calculated by velError jInv + posError jInv
\item[jInv] is calculated using masses and inertias of connected bodies and constraint geometry. 
 \begin{description}
\item[In constraint cases,] it is mass of body (6000).
\item[In contact case,] it varies below mass of body.
\end{description}
\item[Impulse] is impulse applied to body during timestep.
 \begin{description}
\item[In constraint cases,] it is obtained from btJointFeedback structure.
\item[In contact case,] it is obtained by summing applied impulses from active manifolds.
\end{description}
\item[erp] Error reduction parameter (0...1) is used to handle numerical issues e.q. body drifting. 
Setting erp to 1 would in theory eliminate error in one step but in practice value of 0.2 - 0.8 is used in most cases.
\end{description}
\caption{Internal variables used in \bullet\ in constraint solving.}
\label{fig:solVars}
\end{figure}

@@

Six-dof-spring constraint 2 has optional feature to avoid unstability by automatically softening constraint
spring. Feature is activated if current timeStep is overly large for spring-mass system simulation.
The feature is trigged based on the equation below 
\begin{equation} \label{eq:frequencyLimited}
\sqrt{ k /  m_{min}} dt > 0.25 
\end{equation}
where $k$ is elastic spring stiffness between bodies,
$m_{min}$ is smaller of connected masses or inertias and $dt$ is integration timestep.

If this feature is active for elastic-plastic constraint 2 constraint  
it does not report positional error but it allows full plastic capacity to be used for correcting
velocity based error instead of force provided by spring.
In this scenario body behaves as depicted in table \ref{tab:ep2BlockValues} i.e. it does not move at all.

@@

\subsection{Charpy impact test}
In a second numerical example, simulation of Charpy impact test is
used to benchmark approach in more complicated scenario. 
Material is steel and density is 7800 $\frac{kg}{m^{3}}$. Young’s modulus is 200 GPa.
Specimen dimensions are 10x10x55 mm with 2 mm notch in middle which is taken into account in calculations.
Support anvils initially have 40 mm open space between them and their width is 40 mm. 
Hammer is 0.5 m wide and 0.25 m high, thickness is 0.02 m and mass is 19.5 kg.
Hammer has 40 mm draft.
If specimen bends about 1.9 radians (108 degrees) it will go between anvils.
Expected energy loss is product of plastic moment of section, hinge angle needed for specimen to go through supports and 
yield stress of specimen. For hinge angle of 1.9 radians and yield stress of 400 MPa expected energy 
loss is 122 J.

@@

\subsection{Demolisher}
Third numerical examples introduces a demolisher.
Demolisher is collection of possible scenarios that can be added to games to provide ductile joints between
rigid bodies.  Rigid concrete bodies are connected by steel enforcement.
Concrete density is 2000 $\frac{kg}{m^3}$ and steel density is 7800 $\frac{kg}{m^3}$. 
Steel is assumed to handle load. Yield stress of steel is 200 MPa.
Gravity is 10  $\frac{m}{s^2}$. Simulation step is $\frac{1}{60} $ s and 10 
iterations are done for single step. Multiple simulation steps are done for each render if needed.

Simulation can be done in real time on modest hardware, e.g. 
on laptop having 2.0 GHz Intel i3 CPU with integrated graphics. 
Figure \ref{fig:demolisher} shows three screenshots from simulation.

@@

Vehicle is modelled as single box btRaycastVehicle with four driving wheels. Wheels are rendered for visual feedback.
In left picture, initial state is shown. Gate in front is modelled to be so slender that it has visible elastic deflection.
Gate support is modelled as heavy box.
Bridge has rigid ramps and supports at both ends.
Gravity is applied gradually to avoid collapse of bridge at start of simulation.
In middle picture,  bridge is shown in state where car has driven once over bridge without stopping and
other bodies have taken light damage.
In right picture, two joints have been broken and bridge has fallen from support due to bending.

### 2016-10-08 pdocs/thesis/bullet-plasticity.tex
Aki's comments on article and in general

\begin{itemize}
\item Avoid very small and very larger collision shapes
\item Avoid large mass ratios
\end{itemize}

@@

\subsubsection{Breaking of constraints between objects}
For standard \bullet constraints breakingImpulseThreshold (BITh) can be defined. If impulse is larger than set limit constraint is activated. 
This allows object breaking to be simulated with very cheap calculation. 
This method is too simplified for ductile material and developing more precise and realistic methods is main objective of this work.

@@

\subsection{Constraint with zero limits}
In this case btGeneric6DofConstraint is used with high and low limits set to zero. 
Constraint is made breakable by setting breakingImpulseThreshold.
This is common technique to provide breakable objects in games and provides visually accetable results for brittle materials.

\subsection{SpringConstraint}
This shows how btGeneric6DofSpringConstraints act in case like this. 
Specimen can absorb very little elastic energy so rotational spring constants are calculated using plastic state i.e.
$W=bh^2/4$ where h=b-0.002m and yield stress is used instead of Young’s modulus for rotational springs.

@@

\section{Charpy impact test results using \bullet}
\label{sec:bullet-charpy-results}
Results are quite sensitive and obtained in February 2016 using following processes
\begin{itemize}
\item single precision debug build is used
\item sequential impulse constraint solver with 10 iterations is used
\item energy loss values are rounded to nearest 10 joules 
\end{itemize}

@@

\subsection{Results for spring2Constraint}
Results are quite similar as for springConstraint but automatic parameter tuning prevents unstable simulations.
Energy loss is about 40 J.

@@

\subsection{Results for plasticHingeConstraint}
Results are quite similar as for hinge constraint with motors. 
Breaking point can be configured based on rotation angle which makes material behaviour more realistic.
In figure \ref{fig:f6-results} maximum plastic rotation is limited to 1.5 and energy loss is about 95 J.

@@

\subsection{Results for elasticPlasticConstraint}
Results are quite similar as for plasticHingeConstraint as this case is dominated by plastic bending.
Additional logic helps to avoid stablity issues.

\subsection{Results for elasticPlastic2Constraint}
Results are quite similar as for elasticPlasticConstraint and this constraint seems to best candidate for further development.
Table \ref{tab:ep2ts} shows energy loss for few timesteps as example to demonstrate sensitivity of solution.

@@

\subsection{Summary and evaluation of results}
Suggested methods for handling plasticity are well suited for cases where simulation of ductile failure is desired.
It requires only slightly more computational effort than methods that are used for simulation of brittle failures.
Hinge constraint with motors method can be used without introducing any additional code to most existing dynamic engine
as it is based on widely used motor-codes. 
Additional features are quite simple to implement and make spring based constraints even more usable than currently by 
allowing stiff constraints to be used without instability.

>>>>

\begin{itemize}
\item Avoid small and large collision shapes
\item Avoid large mass ratios
\end{itemize}

@@

\subsubsection{Breaking of constraints between bodies}
For standard \bullet constraints breakingImpulseThreshold (BITh) can be defined. If impulse is larger than set limit constraint is activated. 
This allows breaking to be simulated with fast calculation. 
This method is too simplified for ductile material and developing more precise and realistic methods is main objective of this work.

@@

\subsection{Constraint with zero limits}
In this case btGeneric6DofConstraint is used with high and low limits set to zero. 
Constraint is made breakable by setting breakingImpulseThreshold.
This is common technique to provide breakable bodies in games and provides visually accetable results for brittle materials.

\subsection{SpringConstraint}
This shows how btGeneric6DofSpringConstraints act in case like this. 
Specimen can absorb little elastic energy so rotational spring constants are calculated using plastic state i.e.
$W=bh^2/4$ where h=b-0.002m and yield stress is used instead of Young’s modulus for rotational springs.

@@

\section{Charpy impact test results using \bullet}
\label{sec:bullet-charpy-results}
Results are sensitive and they were obtained in February 2016 using following processes
\begin{itemize}
\item single precision debug build is used
\item sequential impulse constraint solver with 10 iterations is used
\item energy loss values are rounded to nearest 10 joules 
\end{itemize}

@@

\subsection{Results for spring2Constraint}
Results are similar as for springConstraint but automatic parameter tuning prevents unstable simulations.
Energy loss is about 40 J.

@@

\subsection{Results for plasticHingeConstraint}
Results are similar as for hinge constraint with motors. 
Breaking point can be configured based on rotation angle which makes material behaviour more realistic.
In figure \ref{fig:f6-results} maximum plastic rotation is limited to 1.5 and energy loss is about 95 J.

@@

\subsection{Results for elasticPlasticConstraint}
Results are similar as for plasticHingeConstraint as this case is dominated by plastic bending.
Additional logic helps to avoid stablity issues.

\subsection{Results for elasticPlastic2Constraint}
Results are similar as for elasticPlasticConstraint and this constraint seems to best candidate for further development.
Table \ref{tab:ep2ts} shows energy loss for few timesteps as example to demonstrate sensitivity of solution.

@@

\subsection{Summary and evaluation of results}
Suggested methods for handling plasticity are well suited for cases where simulation of ductile failure is desired.
It requires only slightly more computational effort than methods that are used for simulation of brittle failures.
Hinge constraint with motors method can be used without introducing any additional code to most existing dynamic engine
as it is based on widely used motor-codes. 
Additional features are simple to implement and make spring based constraints even more usable than currently by 
allowing stiff constraints to be used without instability.
